---
author:
  Michael Braun\
  Edwin L. Cox School of Business\
  Southern Methodist University
bibliography:
- sparseMVN.bib
date: September 26, 2021
title: "An R Package for Multivariate Normal Functions with Sparse
  Covariance and Precision Matrices"
---


```{r, echo=FALSE, cache=FALSE}
suppressPackageStartupMessages(library("sparseMVN"))
suppressPackageStartupMessages(library("Matrix"))
suppressPackageStartupMessages(library("mvtnorm"))
suppressPackageStartupMessages(library("tidyverse"))
suppressPackageStartupMessages(library("kableExtra"))

knitr::opts_chunk$set(prompt=TRUE, cache=FALSE,error=FALSE,
                      echo=FALSE, eval=TRUE,
                      comment="#", collapse=FALSE) #$
options(replace.assign=TRUE, width=77,
        digits=4, xtable.include.rownames=FALSE,
        dplyr.summarise.inform=FALSE)
sanitize <- function(x) x
```

[Matrix]:  https://CRAN.R-project.org/package=Matrix
[sparseMVN]: https://braunm.github.io/sparseMVN
[sparseHessianFD]: https://braunm.github.io/sparseHessianFD
[trustOptim]: https://braunm.github.io/trustOptim
[rmvnorm]: https://rdrr.io/pkg/mvtnorm/man/Mvnorm.html
[dmvnorm]: https://rdrr.io/pkg/mvtnorm/man/Mvnorm.html
[mvtnorm]: https://cran.r-project.org/package=mvtnorm


The [mvtnorm] package [@R_mvtnorm] provides the dmvnorm() function to compute the density of a multivariate normal (MVN) distribution, and the rmvnorm() function to simulate MVN random variables. These functions require the user to supply a full, "dense" covariance matrix; if starting with a precision matrix, the user must first invert it explicitly. This covariance matrix is dense in the sense that, for an $M$-dimensional MVN random variable, all $M^2$ elements are stored, so memory requirements grow quadratically with the size of the problem. Internally, both functions factor the covariance matrix using a Cholesky decomposition, whose complexity is $\mathcal{O}\!\left(M^3\right)$[@GolubVanLoan1996].[^1] This factorization is performed every time the function is called, even if the covariance matrix does not change from call to call. Also, the internal algorithm in [rmvnorm] involves multiplication of a triangular matrix, and for dmvnorm(), involves solving a triangular linear system. Both of these operations are $\mathcal{O}\!\left(M^2\right)$ [@GolubVanLoan1996] on dense matrices. MVN functions in other packages, such as [@R_MASS] and [@R_LaplacesDemon], face similar limitations.[^2] Thus, existing tools for working with the MVN distribution in are not practical for high-dimensional MVN random variables.

However, for many applications the covariance or precision matrix is sparse, meaning that the proportion of nonzero elements is small, relative to the total size of the matrix. The functions in the `sparseMVN` package exploit that sparsity to reduce memory requirements, and to gain computational efficiencies. The `dmvn.sparse()` function computes the MVN density, and the `rmvn.sparse()` function samples from an MVN distribution. Instead of requiring the user to supply a dense covariance matrix, `rmvn.sparse()` and `dmvn.sparse()` accept a pre-computed Cholesky decomposition of either the covariance or precision matrix in a compressed sparse format. This approach has several advantages:

1.  Memory requirements are smaller because only the nonzero elements of the matrix are stored in a compressed sparse format.

2.  Linear algebra algorithms that are optimized for sparse matrices are
    more efficient because they avoid operations on matrix elements that
    are known to be zero.

3.  When the precision matrix is initially available, the user can avoid
    the need to explicitly invert it into a covariance matrix. This
    feature of preserves sparsity, because the inverse of a sparse
    matrix is not necessarily sparse.

4.  The Cholesky factor of the matrix is computed once, before the first
    function call, and is not repeated with subsequent calls (as long as
    the matrix does not change).

The functions in `sparseMVN` rely on sparse matrix classes and functions defined in the [Matrix] package. The user creates the covariance or precision matrix as a sparse, symmetric *dsCMatrix* matrix, and computes the sparse Cholesky factor using the Matrix::Cholesky() function. Other than ensuring that the factor for the covariance or precision matrix is in the correct format, the `sparseMVN` functions behave in much the same way as the corresponding `mvtnorm` functions. Internally, `sparseMVN`  uses standard methods of computing the MVN density and simulating MVN random variables (see Section [1.1](#sec:algorithms){reference-type="ref" reference="sec:algorithms"}). Since a large proportion of elements in the matrix are zero, we need to store only the row and column indices, and the values, of the unique nonzero elements. The efficiency gains in [sparseMVN] come from storing the covariance or precision matrix in a compressed format without explicit zeros, and applying linear algebra routines that are optimized for those sparse matrix structures. [Matrix] calls sparse linear algebra routines that are implemented in the `CHOLMOD` library [@ChenDavis2008; @DavisHager1999; @DavisHager2009].

# Background

Let $x\in\mathbb{R}^{M}$ be a realization of random variable
$X\sim\mathbf{MVN}\!\left(\mu,\Sigma\right)$, where
$\mu\in\mathbb{R}^{M}$ is a vector, $\Sigma\in\mathbb{R}^{M\times M}$ is
a positive-definite covariance matrix, and
$\Sigma^{-1}\in\mathbb{R}^{M\times M}$ is a positive-definite precision
matrix.

The log probability density of $x$ is

$$\begin{aligned}
\log f(x)&=-\frac{1}{2}\left(M \log (2\pi) + \log|\Sigma|
  +z^\top z\right),\quad\text{where}~z^\top z=\left(x-\mu\right)^\top\Sigma^{-1}\left(x-\mu\right)
 \end{aligned}$$

## MVN density computation and random number generation {#sec:algorithms}

The two computationally intensive steps in evaluating $\log f(x)$ are computing $\log|\Sigma|$, and $z^\top z$, *without* explicitly inverting $\Sigma$ or repeating mathematical operations. But once we have $z$, computing the dot product $z^\top z$ is cheap. How one these steps *efficiently* in practice depends on whether the covariance matrix $\Sigma$, or the precision matrix $\Sigma^{-1}$ is available. For both cases, we start by finding a lower triangular matrix root: $\Sigma=LL^\top$ or $\Sigma^{-1}=\Lambda\Lambda^\top$. Since $\Sigma$ and $\Sigma^{-1}$ are positive definite, we will use the Cholesky decomposition, which is the unique matrix root with all positive elements on the diagonal.

With the Cholesky decomposition in hand, we compute the log determinant
of $\Sigma$ by adding the logs of the diagonal elements of the factors.
$$\begin{aligned}
  \label{eq:logDet}
  \log|\Sigma|= \begin{cases}
    \phantom{-}2\sum_{m=1}^M\log L_{mm}&\text{ when $\Sigma$ is given}\\
    -2\sum_{m=1}^M\log \Lambda_{mm}&\text{ when $\Sigma^{-1}$ is given}
    \end{cases}\end{aligned}$$

Having already computed the triangular matrix roots also speeds up the
computation of $z^\top z$. If $\Sigma^{-1}$ is given,
$z=\Lambda^\top(x-\mu)$ can be computed efficiently as the product of an
upper triangular matrix and a vector. When $\Sigma$ is given, we find
$z$ by solving the lower triangular system $Lz=x-\mu$. The subsequent
$z^\top z$ computation is trivially fast.

The algorithm for simulating $X\sim\mathbf{MVN}\!\left(\mu,\Sigma\right)$ also depends on whether $\Sigma$ or $\Sigma^{-1}$ is given. As above, we start by computing the Cholesky decomposition of the given covariance or precision matrix.  Define a random variable $Z\sim\mathbf{MVN}\!\left(0,I_M\right)$, and generate a realization $z$ as a vector of $M$ samples from a standard normal distribution. If $\Sigma$ is given, then evaluate $x=Lz+\mu$. If $\Sigma^{-1}$ is given, then solve for $x$ in the triangular linear system $\Lambda^\top\left(x-\mu\right)=z$. The resulting $x$ is a sample from $\mathbf{MVN}\!\left(\mu,\Sigma\right)$. The following shows that this approach is correct:

$$\begin{aligned}
 \mathbf{E}\!\left(X\right)&=\mathbf{E}\!\left(LZ+\mu\right)=\mathbf{E}\!\left(\Lambda^\top Z+\mu\right)=\mu\\
   \mathbf{cov}\!\left(X\right)&= \mathbf{cov}\!\left(LZ+\mu\right)=\mathbf{E}\!\left(LZZ^\top L^\top\right)=LL^\top=\Sigma\\
     \mathbf{cov}\!\left(X\right)&=\mathbf{cov}\!\left((\Lambda^\top)^{-1}Z+\mu\right)=\mathbf{E}\!\left((\Lambda^\top)^{-1}ZZ^\top\Lambda^{-1}\right)
     =(\Lambda^\top)^{-1}\Lambda^{-1}=(\Lambda\Lambda^\top)^{-1}=\Sigma
 \end{aligned}$$

These algorithms apply when the covariance/precision matrix is either
sparse or dense. When the matrix is dense, the computational complexity
is $\mathcal{O}\!\left(M^3\right)$ for a Cholesky decomposition, and
$\mathcal{O}\!\left(M^2\right)$ for either solving the triangular linear
system or multiplying a triangular matrix by another matrix
[@GolubVanLoan1996]. Thus, the computational cost grows cubically with
$M$ before the decomposition step, and quadratically if the
decomposition has already been completed. Additionally, the storage
requirement for $\Sigma$ (or $\Sigma^{-1}$) grows quadratically with
$M$.

## Sparse matrices in R {#sec:sparse}

The [Matrix] package [@R_Matrix] defines various classes for storing sparse
matrices in compressed formats. The most important class for our
purposes is *dsCMatrix*, which defines a symmetric matrix, with numeric
(double precision) elements, in a column-compressed format. Three
vectors define the underlying matrix: the unique nonzero values (just
one triangle is needed), the indices in the value vector for the first
value in each column, and the indices of the rows in which each value is
located. The storage requirements for a sparse $M\times M$ symmetric
matrix with $V$ unique nonzero elements in one triangle are for $V$
double precision numbers, $V+M+1$ integers, and some metadata. In
contrast, a dense representation of the same matrix stores $M^2$ double
precision values, regardless of symmetry and the number of zeros. If $V$
grows more slowly than $M^2$, the matrix becomes increasingly sparse (a
smaller percentage of elements are nonzero), with greater efficiency
gains from storing the matrix in a compressed sparse format.

### An example

To illustrate how sparse matrices require less memory resources when
compressed than when stored densely, consider the following example,
which borrows heavily from the vignette of the package
[@R_sparseHessianFD].

Suppose we have a dataset of $N$ households, each with $T$ opportunities
to purchase a particular product. Let $y_i$ be the number of times
household $i$ purchases the product, out of the $T$ purchase
opportunities, and let $p_i$ be the probability of purchase. The
heterogeneous parameter $p_i$ is the same for all $T$ opportunities, so
$y_i$ is a binomial random variable.

Let $\beta_i\in\mathbb{R}^{k}$ be a heterogeneous coefficient vector
that is specific to household $i$, such that
$\beta_i=(\beta_{i1},\dotsc,\beta_{ik})$. Similarly,
$w_i\in\mathbb{R}^{k}$ is a vector of household-specific covariates.
Define each $p_i$ such that the log odds of $p_i$ is a linear function
of $\beta_i$ and $w_i$, but does not depend directly on $\beta_j$ and
$w_j$ for another household $j\neq i$: $$\begin{aligned}
  p_i=\frac{\exp(w_i'\beta_i)}{1+\exp(w_i'\beta_i)},~i=1,\dots, N\end{aligned}$$

The coefficient vectors $\beta_i$ are distributed across the population
of households following a MVN distribution with mean
$\mu\in\mathbb{R}^{k}$ and covariance
$\mathbf{A}\in\mathbb{R}^{k\times k}$. Assume that we know $\mathbf{A}$,
but not $\mu$, so we place a multivariate normal prior on $\mu$, with
mean $0$ and covariance $\mathbf{\Omega}\in\mathbb{R}^{k\times k}$.
Thus, the parameter vector $x\in\mathbb{R}^{(N+1)k}$ consists of the
$Nk$ elements in the $N$ $\beta_i$ vectors, and the $k$ elements in
$\mu$.

The log posterior density, ignoring any normalization constants, is
$$\begin{aligned}
  \label{eq:LPD}
  \log \pi(\beta_{1:N},\mu|\mathbf{Y}, \mathbf{W}, \mathbf{A},\mathbf{\Omega})=\sum_{i=1}^N\left(p_i^{y_i}(1-p_i)^{T-y_i}
  -\frac{1}{2}\left(\beta_i-\mu\right)^\top\mathbf{A}^{-1}\left(\beta_i-\mu\right)\right)
-\frac{1}{2}\mu^\top\mathbf{\Omega}^{-1}\mu\end{aligned}$$

Because one element of $\beta_i$ can be correlated with another element
of $\beta_i$ (for the same unit), we allow for the cross-partials
between elements of $\beta_i$ for any $i$ to be nonzero. Also, because
the mean of each $\beta_i$ depends on $\mu$, the cross-partials between
$\mu$ and any $\beta_i$ can be nonzero. However, since the $\beta_i$ and
$\beta_j$ are independent samples, and the $y_i$ are conditionally
independent, the cross-partial derivatives between an element of
$\beta_i$ and any element of any $\beta_j$ for $j\neq i$, must be zero.
When $N$ is much greater than $k$, there will be many more zero
cross-partial derivatives than nonzero, and the Hessian of the log
posterior density will be sparse.


```{r, echo=FALSE}
N <- 5
k <- 2
p <- k ## dimension of mu
nv1 <- N*k+p
nels1 <- nv1^2
nnz1 <- N*k^2 + 2*p*N*k + p^2
nnz1LT <- N*k*(k+1)/2  + p*N*k + p*(p+1)/2
Q <- 1000
nv2 <- Q*k+p
nels2 <- nv2^2
nnz2 <- Q*k^2 + 2*p*Q*k + p^2
nnz2LT <- Q*k*(k+1)/2 + p*Q*k + p*(p+1)/2
options(scipen=999)
```





The sparsity pattern depends on how the variables are ordered. One such
ordering is to group all of the coefficients in the $\beta_i$ for each
unit together, and place $\mu$ at the end. $$\begin{aligned}
\beta_{11},\dotsc,\beta_{1k},\beta_{21},\dotsc,\beta_{2k},~\dotsc~,\beta_{N1},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k\end{aligned}$$
In this case, the Hessian has a "block-arrow" pattern. Figure BLOCKARROW  illustrates this pattern for $N=`r scales::comma(N)`$
and $k=`r k`$ (`r nv1` total variables).

```{r blackarrow, results="asis"}
Mat1a <- as(kronecker(diag(N), matrix(1, k, k)),"sparseMatrix")
Mat1a <- rbind(Mat1a, Matrix(1, p, N*k))
Mat1a <- cbind(Mat1a, Matrix(1, k*N+p, p))

tmp_a <- as.matrix(Mat1a)
colnames(tmp_a) <- letters[1:NCOL(tmp_a)]


tmp_a %>%
  as_tibble(.name_repair="minimal") %>%
  mutate(across(colnames(tmp_a), ~ifelse(.x == 1, "|", "."))) %>%
  kable(col.names=NULL, caption="A block-arrow pattern") %>%
  kable_paper()

```



Another possibility is to group coefficients for each covariate
together. $$\begin{aligned}
\beta_{11},\dotsc,\beta_{N1},\beta_{12},\dotsc,\beta_{N2},~\dotsc~,\beta_{1k},\dotsc,\beta_{Nk},\mu_1,\dotsc,\mu_k\end{aligned}$$
Now the Hessian has an \"banded\" sparsity pattern, as in
Figure BANDED.



```{r, banded, echo=FALSE,results="asis"}
Mat1b <- kronecker(Matrix(1, k, k), diag(N))
Mat1b <- rbind(Mat1b, Matrix(1, p, N * k))
Mat1b <- cbind(Mat1b, Matrix(1, k*N+p, p))
tmp_b <- as.matrix(Mat1b)
colnames(tmp_b) <- letters[1:NCOL(tmp_b)]


tmp_b  %>%
  as_tibble(.name_repair="minimal") %>%
  mutate(across(colnames(tmp_b), ~ifelse(.x == 1, "|", "."))) %>%
  kable(col.names=NULL, caption="A banded sparsity pattern") %>%
  kable_paper()

```


```{r, echo=FALSE, results="hide"}
Mat2 <- as(kronecker(diag(Q),matrix(1,k,k)),"lMatrix") %>%
    rbind(Matrix(TRUE,p,Q*k)) %>%
    cbind(Matrix(TRUE, k*Q+p, p)) %>%
    as("dgCMatrix") %>%
    as("symmetricMatrix")
A2 <- as(Mat2,"matrix")
```



In both cases, the number of nonzeros is the same.  There are `r nels1` elements in this symmetric matrix.  If the
matrix is stored in the standard \pkg{base} \proglang{R} dense
format, memory is reserved for all `r nels1` values, even though only `r nnz1` values are
nonzero, and only `r nnz1LT` values are unique. For larger
matrices, the reduction in memory requirements by storing the matrix
in a sparse format can be substantial.\footnote{Because sparse matrix
  structures store row and column indices of the nonzero values, they
  may use more memory than dense storage if the total number of
  elements is small}  If $N=`r scales::comma(Q)`$ and $k=`r k`$, then
$M=$`r scales::comma(nv2)`, with more than $`r floor(nels2/10^6)`$ million
elements in the Hessian.  However, only `r scales::comma(nnz2)` of those elements are
nonzero, with `r scales::comma(nnz2LT)` unique values in the lower
triangle.  The dense matrix requires `r format(object.size(A2),units='Mb')` of RAM,
while a sparse symmetric matrix of the \class{dsCMatrix} class requires only `r format(object.size(Mat2),units='Kb')`.



This example is relevant because, when evaluated at the posterior mode,
the Hessian matrix of the log posterior is the MVN precision matrix
$\Sigma^{-1}$ of a MVN approximation to the posterior distribution of
$\left(\beta_{1:N},\mu\right)$. If we were to simulate from this MVN
using , or evaluate MVN densities using , we would need to invert
$\Sigma^{-1}$ to $\Sigma$ first, and store it as a dense matrix.
Internally, and invoke dense linear algebra routines, including matrix
factorization.

# Using the sparseMVN package

The signatures of the key sparseMVN functions are

```{r, eval=FALSE, prompt=FALSE}
rmvn.sparse(n, mu, CH, prec=TRUE, log=TRUE)
dmvn.sparse(x, mu, CH, prec=TRUE, log=TRUE)
```





The rmvn.sparse() function returns a matrix $x$ with $n$ rows and `lenght(mu)` columns. `dmvn.sparse()` returns a
vector of length n: densities if `log=FALSE` , and log densities if `log=TRUE`.

| | |
|-----|-----|
| **x** |A numeric matrix. Each row is an MVN sample.|
| **mu** |A numeric vector. The mean of the MVN random variable.|
| **CH** |Either a *dCHMsimpl* or *dCHMsuper* object representing the Cholesky decomposition of the covariance/precision matrix.|
| **prec** |Logical value that identifies CH as the Cholesky decomposition of either a covariance ($\Sigma$, ) or precision($\Sigma^{-1}$, ) matrix.|
| **n** |Number of random samples to be generated.|
| **log** |If `log=TRUE`, the log density is returned.|

Table [1](#tab:args){reference-type="ref" reference="tab:args"} describes the function arguments. These functions do require the user to compute the Cholesky decomposition `CH` beforehand, but this needs to be done only once (as long as $\Sigma$ or $\Sigma^{-1}$ does not change). `CH` should be computed using  Matrix::Cholesky(), whose first argument is a sparse symmetric matrix stored as a *dsCMatrix* object. As far as we know, there is no particular need to deviate from the defaults of the remaining arguments. If Matrix::Cholesky() uses a fill-reducing permutation to compute `CH` , the sparseMVN functions will handle that directly, with no additional user intervention required. The chol() function in base R should not be used.

## An example {#sec:example}

Suppose we want to generate samples from an MVN approximation to the
posterior distribution of our example model from Section
[1.2](#sec:sparse){reference-type="ref" reference="sec:sparse"}.
sparseMVN includes functions to simulate data for the example (binary.sim()), and to compute
the log posterior density (binary.f()), gradient (binary.grad()), and Hessian (binary.hess()).  trustOptim::trust.optim() [@R_trustOptim] is a nonlinear optimizer that estimates
the curvature of the objective function using a sparse Hessian.


```{r, results='hide'}
D <- sparseMVN::binary.sim(N=50, k=2, T=50)
priors <- list(inv.A=diag(2), inv.Omega=diag(2))
start <- rep(c(-1,1),51)
opt <- trustOptim::trust.optim(start,
                               fn=sparseMVN::binary.f,
                               gr=sparseMVN::binary.grad,
                               hs=sparseMVN::binary.hess,
                               data=D, priors=priors,
                               method="Sparse",
                               control=list(function.scale.factor=-1))
```







The calls to trustOptim::trust.optim()  return the posterior mode, and the Hessian evaluated at the
mode. These results serve as the mean and the negative precision of the
MVN approximation to the posterior distribution of the model.


```{r}
R <- 100
pm <- opt[["solution"]]
H <- -opt[["hessian"]]
CH <- Cholesky(H)
```



We can then sample from the posterior using an MVN approximation, and
compute the MVN log density for each sample.

```r
samples <- rmvn.sparse(R, pm, CH, prec=TRUE)
logf <- dmvn.sparse(samples, pm, CH, prec=TRUE)
```


The ability to accept a precision matrix, rather than having to invert
it to a covariance matrix, is a valuable feature of sparseMVN . This is because
the inverse of a sparse matrix is not necessarily sparse. In the
following chunk, we invert the Hessian, and drop zero values to maintain
any remaining sparseness. Note that there are `r scales::comma(102^2)`  total elements in
the Hessian.

```r
Matrix::nnzero(H)
Hinv <- drop0(solve(H))
Matrix::nnzero(Hinv)
```



Nevertheless, we should check that the log densities from dmvn.sparse() correspond to
those that we would get from mvtnorm::dmvnorm().


```r
logf_dense <- dmvnorm(samples, pm, as.matrix(Hinv), log=TRUE)
all.equal(logf, logf_dense)
```

```{r, echo=FALSE}
load("runtimes.Rdata")
```

```{r, echo=FALSE}
tab1 <- filter(runtimes, stat %in% c("density","rand")) %>%
    group_by(N, k, stat, pattern, type) %>%
    summarize(mean_ms=mean(time/1000000),
              sd_ms=sd(time/1000000)) %>%
    tidyr::gather(time, value, c(mean_ms, sd_ms)) %>%
    reshape2::dcast(N+k+stat+time~pattern+type)


tab2 <- filter(runtimes, stat %in% c("chol","solve")) %>%
    group_by(N, k, stat, pattern, type) %>%
    summarize(mean_ms=mean(time/1000000),
              sd_ms=sd(time/1000000)) %>%
    tidyr::gather(time, value, c(mean_ms, sd_ms)) %>%
    reshape2::dcast(N+k+time~stat+pattern)
```



# Timing

In this section we show the efficiency gains from sparseMVN by comparing the run times between rmvn.sparse() and [rmvnorm], and between dmvn.sparse() and [dmvnorm] . In these tests, we construct covariance/precision matrices with the same structure as the Hessian of the log posterior density of the example model in Section [2.1](#sec:example){reference-type="ref" reference="sec:example"}.  Parameters are ordered such that the matrix has a [block-arrow pattern](#blockarrow).

<!-- as in Figure [\[fig:blockarrow\]](#fig:blockarrow){reference-type="ref" reference="fig:blockarrow"}.  -->
as in Figure [\[fig:blockarrow\]](#fig:blockarrow)

The size and sparsity of the test matrices vary through manipulation of the number of blocks ($N$), the size of each block ($k$), and the number of rows/columns in the margin (also $k$). Each test matrix has $(N+1)k$ rows and columns.  Table [2](#tab:cases){reference-type="ref" reference="tab:cases"} summarizes the case conditions.


```{r, echo=FALSE, fig.cap="Cases for timing comparison. $N$ and $k$ refer, respectively, to the number of blocks in the block-arrow structure (analogous to heterogeneous units in the  binary choice example), and the size of each block.  The total number of variables is $M=(N+1)k$, and the total number of elements in the matrix is $M^2$. The three rightmost columns present the number of nonzeros in the full matrix and lower triangle, and the sparsity (proportion of matrix elements that are nonzero)."}

## tmp <- c("\\multirow{8}{*}{k=2}",rep(NA,7),
##          "\\multirow{8}{*}{k=4}",rep(NA,7))
cases %>%
  select(k,N, nvars,nels,
         nnz, nnzLT, pct.nnz) %>%
  kable(digits=c(rep(0,6),3), format.args=list(big.mark=","),
        col.names=c('k', 'N', 'variables', 'elements', 'full', 'lower tri', 'sparsity')) %>%
  kable_paper()  %>%
   add_header_above(c(" "=4, "nonzeros"=3))

  ## xtable::xtable(digits=c(rep(0,7),3)) %>%
##     print(include.rownames=FALSE,only.contents=TRUE,
##           include.colnames=FALSE,
##           floating=FALSE, sanitize.colnames.function=sanitize,
##           sanitize.text.function=sanitize,
##           hline.after=8,
##
## -
```


```{r, echo=FALSE}
tmp <- filter(tab1, N==min(tab1[['N']]) & k==min(tab1[['k']])
              & time=="mean_ms" & stat=="density")
sm <- with(tmp, c(dense_cov,sparse_cov))
```


Figure~\ref{fig:densRand} compares mean run times to compute 1,000 MVN densities, and generate 1,000
MVN samples, using rmvn.sparse() and dmvn.sparse() from sparseMVN, and
mvtnorm::dmvnorm() and mvtnorm::rmvnorm().  Times were collected over 200 replications on a
  2013-vintage Mac Pro with a 12-core 2.7 GHz Intel Xeon E5 processor
  and 64 GB of RAM.
  The times for \pkg{mvtnorm} are faster than
  \pkg{sparseMVN} for low dimensional cases ($N\leq 50$), but grow
  quadratically in the number of variables.\footnote{As an example, in
    the $N=`r min(tab1[['N']])`$, $k=`r min(tab1[['k']])`$ case, the mean time to compute 1,000 MVN
    densities is `r format(sm[1],digits=2)` ms using [dmvnorm], but more
    than `r format(sm[2],digits=2)` ms using dmvn.sparse().}  This is because the number
  of elements stored in a dense covariance matrix grows quadratically
  with the number of variables. In this example,
  storage and computation requirements for the sparse matrix grow linearly with the number
  of variables, so the `sparseMVN` run times grow linearly as
  well \citep[sec.~4]{BraunDamien2016}.  The comparative advantage of `sparseMVN` increases with
  the sparsity of the covariance matrix.[^5]



```{r, echo=FALSE, fig.height=5, fig.cap="Mean computation time for simulating 1,000 MVN samples, and computing 1,000 MVN densities, averaged over 200 replications. Densities were computed using [dmvnorm] and dmvn.sparse(), while random samples were generated using [rmvnorm] and rmvn.sparse()."}
DF1 <- filter(tab1, time=="mean_ms") %>%
    mutate(stat=plyr::revalue(stat, c(density="density", rand="random"))) %>%
    rename(dense=dense_cov, sparse=sparse_cov) %>%
  tidyr::gather(pattern, value, c(dense, sparse))

P1 <- ggplot(DF1, aes(x=N, y=value, color=pattern, shape=pattern, linetype=pattern)) +
  geom_line() +
  geom_point(size=2) +
  scale_x_continuous("Number of blocks (N)") +
  scale_y_continuous("Computation time (milliseconds)", labels=scales::comma) +
  scale_color_manual("Pattern",values=c(dense='red', sparse='blue')) +
  scale_shape("Pattern") +
  scale_linetype("Pattern") +
  facet_grid(stat~k, scales="free_y", labeller=label_bquote(cols = k==.(k))) +
  theme_bw() +
  theme(strip.background=element_rect(fill='white'))

print(P1)

```

The sparseMVN functions always require a sparse Cholesky decomposition of the covariance or precision matrix, and the `mvtnorm` functions require a dense precision matrix to be inverted into a dense covariance matrix. Figure [2](#fig:cholSolve){reference-type="ref" reference="fig:cholSolve"} compares the computation times of these preparatory steps. There are three cases to consider: inverting a dense matrix using the solve() function, decomposing a sparse matrix using Matrix::Cholesky(), and decomposing a dense matrix using chol(). Applying chol() to a dense function is not a required operation in advance of calling [dmvnorm] or [rmvnorm], but those functions will invoke some kind of decomposition internally. We include it in our comparison because it comprises a substantial part of the computation time. The decomposition and inversion operations on the dense matrices grow cubically as the size of the matrix increases. The sparse Cholesky decomposition time is negligible. For example, the mean run time for the $N=500$, $k=4$ case is about 0.39 ms.

fig.cap="Computation time for Cholesky decomposition of sparse and dense matrices, and inversion of dense matrices."


```{r, DF2, echo=FALSE, fig.keep='all'}
DF2 <- filter(tab2, time=="mean_ms") %>%
    rename(`dense inversion`=solve_dense,
           `dense Cholesky`=chol_dense, `sparse Cholesky`=chol_sparse) %>%
  tidyr::gather(pattern, value, c(`dense inversion`,`sparse Cholesky`,`dense Cholesky`))

P2 <- ggplot(DF2, aes(x=N, y=value, color=pattern, shape=pattern, linetype=pattern)) +
  geom_line() +
  geom_point(size=2) +
  scale_x_continuous("Number of blocks (N)") +
  scale_y_continuous("Computation time (milliseconds)", labels=scales::comma) +
  scale_color_manual("Pattern/Operation",
                         values=c(`dense Cholesky`='red', `sparse Cholesky`='blue',
                                  `dense inversion`='black')) %>%
  scale_shape("Pattern/Operation") +
  scale_linetype("Pattern/Operation") +
  facet_grid(.~k, scales="free_y", labeller=label_bquote(cols = k==.(k))) +
  theme_bw() +
  theme(strip.background=element_rect(fill='white'))

print(P2)

```






[^1]: has options for eigen and singular value decompositions. These are
    both $\mathcal{O}\!\left(M^3\right)$ as well.

[^2]: LaplacesDemon does offer options for the user to supply pre-factored covariance and precision matrices. This avoids repeated calls to the $\mathcal{O}\!\left(M^3\right)$ factorization step, but not the $\mathcal{O}\!\left(M^2\right)$ matrix multiplication and linear system solution steps.

[^3]: Because sparse matrix structures store row and column indices of
    the nonzero values, they may use more memory than dense storage if
    the total number of elements is small

[^4]: As an example, in the $N=10$, $k=2$ case, the mean time to compute
    1,000 MVN densities is 1.1 ms using , but more than 3.7 ms using .

[^5]: Across all cases there was hardly any difference in the run times
    when providing the precision matrix instead of the covariance.

# References
